Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-25 paper/paper.tex
Add the discussion about parallel reduction vs lo/wh reduction

In terms of our proof technique, we close over our semantically valid types
and terms in \rref{I-Red, I-Bool} using the non-deterministic parallel
reduction relation, while \citet{decagda,nbeincoq} employ a deterministic weak
head reduction relation. Our use of a non-deterministic reduction strategy
means that we need confluence to prove the functionality of our logical
relation. However, the benefit is that it immediately gives us a semantic
justification of the conversion rule (Lemma~\ref{lemma:logrelcoherence}).
% In particular, we obtain the confluence
% result of reduction at a very early stage before we even define our
% logical relation.

@@

% Finally, it is worth pointing out that all systems we have discussed
% so far builds a relational model for their logical relation.

% Recall
% that our logical relation takes the form $[[InterpR i A S]]$ where
% $[[S]]$ is a set of terms. In a relational model, each type is
% interpreted as a partial equivalence relation over terms rather than a
% set. A relational model is necessary for an extensional type theory
% such as Nuprl, though \citet{nbeincoq,decagda,martin-lof-a-la-coq} all
% use a relational model either to justify the function $\eta$-law or to
% derive $\Pi$-injectivity. In Section~\ref{sec:extension}, our $\eta$
% law for function is baked into the untyped reduction
% relation and therefore a relational model is not needed.
% \scw{would it be difficult to extend your proof to be a PER?}

% The dependently typed systems we have discussed so far also vary
% greatly in expressiveness. For example, the Calculus of
% Constructions~\citep{CoC} and MLTT without universe levels are unable
% to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
% identity types. % \citet{nbeincoq} lacks $\eta$-law for functions.
% Among
% the mechanized systems we have discussed so far, only
% \citet{anand2014towards} has a full universe hierarchy.
% Our object language \lang{}, while small, has a decent coverage of
% features commonly seen in dependently typed languages, including
% dependent pattern matching, dependent function types, a full universe
% hierarchy, an intensional identity type. We hope features such as
% type-directed reduction (necessary for unit $\eta$-law) and
% impredicative sorts can be implemented in a similarly streamlined
% fashion and we will leave those extensions as part of our future work.

% To end this section, we
% \scw{should this be a new subsection?}
% The final question we want to address is: why is our proof,
% even with the extension from Section~\ref{sec:extension}, so much
% shorter than the proofs from \citet{decagda, nbeincoq,
%   martin-lof-a-la-coq}?

% This contrasts our development with
% \citet{martin-lof-a-la-coq}, where the irrelevance property turns out to be the most
% difficult property to prove due to the complexity of the equality
% judgment used to define the logical relation. In our system, the
% irrelevance property (Corollary~\ref{lemma:logrelcoherence}) is a
% direct consequence of Lemma~\ref{lemma:interppreservation},\scw{refer to
% lemma name} which is
% straightforward after we prove the results about parallel
% reduction in Section~\ref{sec:spec}. In other words, we keep the
% the logical relation itself minimal at the cost of extra lemmas that can be proven
% through syntactic means independently from the logical relation.
% For example, as commented near the end of Section~\ref{sec:extension},
% our normalization property (Corollary~\ref{corollary:exbetanf})
% % can induce a decision procedure for type conversion by reducing two
% % terms to $\beta\eta$-normal form and check for their syntactic
% % equivalence. This decision procedure would be inefficient compared to
% % NbE or the algorithm from \citet{martin-lof-a-la-coq} that reduces terms
% % to weak head normal and checks for the equivalence of subterms
% % recursively
% does not immediately induce an efficient decision procedure, but it is
% possible to recover an efficient algorithm by reasoning about untyped
% lambda terms.

% the concision of
% our proof can be largely attributed to the fact that we bake into the
% logical relation only the minimum amount of information that is the
% most conducive to the proof of the metatheoretic result.

% Another key to our proof is the very early establishment of the confluence
% property of the paralell reduction relation
% (Lemma~\ref{lemma:pardiamond}). This property is used in
% Lemma~\ref{lemma:interppreservation} to show that the interpretation
% of a type is preserved under evaluation. That is, if $[[InterpR i A S]]$ and $[[A => B]]$, then we also have $[[InterpR i B S]]$.
% In a system with typed directed reduction, the confluence result is
% harder to establish since confluence requires subject reduction, which
% circularly depends on $\Pi$-injectivity, a consequence of
% confluence~\citep{siles2012pure}. There are different workarounds to
% this problem. \citet{siles2012pure} proposes a syntactic approach by
% defining a type system where the confluence result is
% directly provable and later show that the system of interest is
% equivalent to the system with the confluence property.
% \citet{decagda} uses a relational model and the relational counterpart
% of Lemma~\ref{lemma:extwn} to derive $\Pi$-injectivity. The relational
% model is parameterized by a typed indexed equivalence relation and
% is reused to prove derive different properties in their
% development. Another perhaps simpler approach is to simply
% extend the judgmental equality with rules about
% $\Pi$-injectivity~\citep{weirich:systemd}. This allows subject
% reduction to be proven independently from confluence and
% $\Pi$-injectivity can later be shown to be admissible. We find the
% approach from \citep{weirich:systemd} the most lightweight as it
% allows us to derive confluence early on even in systems with typed
% conversion. \scw{Maybe mention my $\eta$-equivalence paper here somewhere?}

% In cases where the judgmental equality does not leverage type information heavily
% (e.g. $\eta$-law for the unit type), it might be possible to show that
% $\Gamma \vdash a \equiv b : A$ is equivalent to the conjunction of
% $[[G |- a : A]]$, $[[G |- b : A ]]$, and $[[a <=> b]]$. If such a
% property holds (we believe this is likely true for both
% \citet{decagda} and \citet{martin-lof-a-la-coq}), we can reuse our existing logical relation in
% terms of untyped parallel reduction to show the decidability of type
% conversion. Otherwise, to fully model the system, we need to bake
% typing information into our logical relation and we need to redefine
% our logical relation in Kripke-style so we can relax the scoping
% constraint in order to prove adequacy, though a relational model would
% still be unnecessary as long as the $\eta$ laws are baked into the
% typed directed relation. \scw{not sure I understand this last bit}

% Since confluence plays such as key role in our proof, we are uncertain
% how to generalize our technique to systems where confluence does not
% hold, and we will investigate those scenarios in our future work.
% \scw{What systems are you thinking of? Or do you mean more generally
% that this technique only works for languages where the definition of
% equality is based on joinability of parallel reduction?}

% We are able to avoid the complexivity of a Kripke-style logical relation
% by defining our logical relation in terms of untyped parallel
% reduction, which does not impose any scoping constraint on the
% terms.

% Why is it that the justification of $\eta$ law
% requires a Kripke-style relational model in all three developments
% but not ours? Are we ``cheating'' by defining \lang{}'s equational
% theory using an untyped conversion rule? Instead of answering the
% questions one by one, we clarify the design of our type system and our
% proof technique so the answers to those questions are self-evident.

% The key to the shortness of

>>>>

On the other hand, our definition of our logical relation does
contribute to a more concise proof.
In \rref{I-Red, I-Bool}, we choose parallel reduction, a full
reduction relation, to close over our semantic interpretation of types
and terms. Parallel reduction is non-deterministic, but it satisfies
useful structural properties such as congruence
(Lemma~\ref{lemma:parcong}) and the diamond property
(Lemma~\ref{lemma:pardiamond}). We pay the price of using a
non-deterministic reduction relation when we want to prove that our
logical relation is a partial function; because of \rref{I-Red}, we
can have $[[A => B0]]$ and $[[A => B1]]$, where $[[B0]]$ and $[[B1]]$
each have their separate interpretations that we have to prove to be
equal. Fortunately, this complexity is easily reconciled by the
diamond property, which is easy to derive syntactically.
\citet{decagda,nbeincoq} employ a deterministic weak
head reduction relation. The use of a deterministic reduction relation
makes the functionality of a logical relation trivial to prove, but
the downside is that weak head reduction (denoted by $\leadsto^{wh}$)
or leftmost-outermost reduction (denoted by $\leadsto^{lo}$) does
not satisfy even the weaker substitution property
(Lemma~\ref{lemma:parsubst}). That is, given a term $[[A]]$
and $a \leadsto^{lo} b$, we do not necessarily have $[[A {a}]]
\leadsto^{lo} [[A {b}]]$ since the variable might appear to the
right of some redex. Eventually, one has to prove how the
deterministic reduction relates to a full non-deterministic reduction
relation to prove the fundamental theorem. In our setup, this would amount
to proving the factorization theorem of the deterministic reduction
relation with respect to full non-deterministic reduction. Instead,
using parallel reduction in \rref{I-Red} allows us to delay the
factorization property until we need to extract a normalization
algorithm from our fundamental theorem (Section~\ref{sec:conversionalgo}).

% In terms of our proof technique, the choice of parallel reduction, a full reduction
% relation, to close over our semantic interpretation in \rref{I-Red,I-Bool} has a non-negligible effect on the
% size of our development.

% we close over our semantically valid types
% and terms in \rref{I-Red, I-Bool} using the non-deterministic parallel
% reduction relation, while \citet{decagda,nbeincoq} employ a deterministic weak
% head reduction relation. Our use of a non-deterministic reduction strategy
% means that we need confluence to prove the functionality of our logical
% relation. However, the benefit is that it immediately gives us a semantic
% justification of the conversion rule (Lemma~\ref{lemma:logrelcoherence}).
% In particular, we obtain the confluence
% result of reduction at a very early stage before we even define our
% logical relation.

@@
